---
title: "ANOVA shinyapp funcs"
output:
  html_document:
    df_print: paged
---

This below is an ANOVA implementation in R.

The shinyapp will be simple and tentatively has the following interface.

### UI design

* Main UI elements are:
  - file input field for data as csv
  - select metric Y variable(s) (variable selection enabled for multiple variables)
  - select grouping X variable(s) (variable selection for multiple variables)
  - select factor X variables (variable selection for multiple variables)
  
* the main output tabs will be as:
  - Overview tab (with example datasets - telecom churn and mtcars)
  - Data summary tab (view top 15 odd obs)
  - ANOVA results tab
  - Plots tab
  
Plan is to have 1 big function handle most of this, using IF statements.

So let's get started.

```{r setup_chunk}

suppressPackageStartupMessages({
  require(tidyverse)
  require(car)
  require(psych)
  })

```

### Data Prep

Function below requires data in a particular format, viz. `data.frame(y, X)` where y and X have been chosen by user. Can restrict max rows to 3k or 5k as needed.

Below I demo data prep for the inbuilt `mtcars` dataset. 

```{r ui_data}

## --- UI for user-input needed below
data = mtcars # input df is renamed 'data'
y = data[,c(1,4)] # user selects Y variable(s), e.g., 'mpg', 'hp' here
X = data[,c(2,6,9)] # user selects X variable(s), e.g.'cyl', 'wt', 'am'
factor_vars = data[, c(2,9)] # user selects factor variables in X

# ensure selected factor_vars are indeed factors.
for (i0 in 1:ncol(X)){  
    if (colnames(X)[i0] %in% colnames(factor_vars)){ X[,i0] = as.factor(X[,i0])}  
    }

head(X)

# show tabular in summary tab below the DF slice
if (ncol(factor_vars) > 1) {xref_tbl = table(factor_vars)}; 
xref_tbl  # display this

```

## The ANOVA grand func

Below I define an `anova_func` that takes care of 3 cases:

* 1-way anova wherein there's one metric Y and one factor X

* ANCOVA wherein there's one metric Y and multiple Xs

* MANOVA wherein we have multiple Ys and multiple Xs

There are 3 results of interest, common from each of the above cases.

* the summary of the anova fit to be shown in ANOVA results tab
* the summary of an OLS alternative fit, to also be shown in the ANOVA results tab
* a plot of group means of the metric Y against groups in factor Xs, for plots tab

Func is below, will walk through indiv cases later.

```{r}

## --- helper funcs
ggplot_aov <- function(y, x, y_name, x_name){
	p = ggplot(data.frame(y,x), aes(x=x, y=y)) +
		geom_boxplot() + 
		labs(x=x_name, y=y_name, color=x_name)  
	return(p)
}

## -- ANOVA grand func

anova_func <- function(y, X, factor_vars){  # These are user-selected from UI

  # summary, summary.lm and plot are the 3 common outputs to be stored
  output_store_list = vector(mode="list", length=3) 

  if (ncol(y) == 1){
  
    if (ncol(X) == 1){
      cat('\n\n ----- Single Y and X: Running 1-way ANOVA -----\n\n')
      y = as.matrix(y)
      fit_1way = aov(y ~ ., data = X)
      y_name = colnames(y); x_name = colnames(X)
      summary(fit_1way)  # show this as anova output
      summary.lm(fit_1way)  # show this as OLS output
      
      # show this plot in plots tab
      p = ggplot_aov(y, X, y_name, x_name)
      p  # display this
  
      # use output_store_list obj below for output tabs
      output_store_list[[1]] = summary(fit_1way)
      output_store_list[[2]] = summary.lm(fit_1way)
      output_store_list[[3]] = ggplot_aov(y, X, y_name, x_name)
      
      } else { # i.e., if ncol(X) >1
        cat('\n\n ----- Single Y, many Xs: Running ANCOVA ------\n\n')
        #y = as.matrix(y); head(y)
        fit_ancova = aov(as.matrix(y) ~ ., data=X)
        summary.lm(fit_ancova)  # show this as OLS result
        
        a0 = car::Anova(fit_ancova, type="III") # use this one for ancova
        a0_df = as.data.frame(a0)
        colsum0 = sum(a0$`Sum Sq`)
        a0_df$percent_var_expl = 100*a0$`Sum Sq`/colsum0 
        a0_df # show as ancova output
        
        # show below plots in plots tab
        n1 = ncol(factor_vars); n1
        plots_store_list = vector(mode="list", length=n1)          
        for (i0 in 1:n1){
          x = as.factor(factor_vars[,i0]); 
          x_name = colnames(factor_vars)[i0]; x_name
          y_name = colnames(y)[1]; y_name
          plots_store_list[[i0]] = ggplot_aov(y, x, y_name, x_name)
          plots_store_list[[i0]] # displaying the stored plot
        } # i0 loop ends
  
      # use output_store_list obj below for output tabs
      output_store_list[[1]] = a0_df
      output_store_list[[2]] = summary.lm(fit_ancova)
      output_store_list[[3]] = plots_store_list      
        
      } # if-else loop ends
  
    } else {  # i.e., if ncol(Y) >1
    
      cat('\n\n ----- Multiple Ys: Running MANOVA ------\n\n')
      fit_manova = manova(as.matrix(y) ~., data=X)
      summary(fit_manova) # show result

      # plot boxplots for manova
      n1 = ncol(factor_vars); n1
      n2 = ncol(y); n2
      plot_store_manova = vector(mode="list", length=n1*n2)
      i0 = 0
      for (i1 in 1:n2){
          y1 = y[,i1]; y_name = colnames(y)[i1]
          for (i2 in 1:n1){
              x = as.factor(factor_vars[,i2])
              x_name = colnames(factor_vars)[i2]
              plot_store_manova[[i0 + 1]] = ggplot_aov(y1, x, y_name, x_name)
              i0 = i0 + 1        } # i2 loop ends
        
        } # i1 loop ends
      
      # plot_store_manova[[2]] # check an example of the stored plots for display
  
      # use output_store_list obj below for output tabs
      output_store_list[[1]] = summary(fit_manova)
      output_store_list[[2]] = summary.lm(fit_manova)
      output_store_list[[3]] = plot_store_manova      
      
      } # if-else loop ends
  
   return(output_store_list) 
   } # func ends

```

## Testing 1-way anova on mtcars

Below is a test-drive on mtcars. Let's try 1way anova first (1 metric Y and 1 factor X). Behold.
 
```{r mtcars_1way}

data = mtcars; # head(mtcars)

y = as.data.frame(data[,1])
X = as.data.frame(as.factor(data[,2])) # must be a factor in 1way anova
colnames(y) = colnames(data)[1] # ensure colname is ame as in dataset
colnames(X) = colnames(data)[2]
factor_vars = X

system.time({
  output_list_mtcars = anova_func(y, X, factor_vars)
})

 output_list_mtcars[[1]]
 output_list_mtcars[[2]]
 output_list_mtcars[[3]] # shows some error but works when run line by line

```
 
## Testing ANCOVA on mtcars

Am taking 1 metric Y ('mpg') and 3 Xs ('cyl', 'am', 'wt') of which 2 are factor and 1 is metric.

Idea is to run ANCOVA and see the results.

```{r ancova_mtcars}

data = mtcars; # head(mtcars)

y = as.data.frame(data[,1])
X = as.data.frame(data[,c(2,6,9)]) 
factor_vars = as.data.frame(data[, c(2,9)])

# ensure colname is same as in dataset
if (ncol(y)==1) {colnames(y) = colnames(data)[1]} 
if (ncol(X)==1) {colnames(X) = colnames(data)[2]}
if (ncol(factor_vars)==1) {colnames(factor_vars) = colnames(data)[2]}

# ensure selected factor_vars are indeed factors.
for (i0 in 1:ncol(X)){  
    if (colnames(X)[i0] %in% colnames(factor_vars)){ X[,i0] = as.factor(X[,i0])}  
    }

# head(X)

system.time({
  output_list_mtcars = anova_func(y, X, factor_vars)
})

output_list_mtcars[[1]]
output_list_mtcars[[2]]
#output_list_mtcars[[3]]

```



Below is a test-drive on the iris dataset for a case with 2 Y variables and 1 X variable

```{r iris_test_drive}
## test drive
data = iris
y = data[,c(1,3)]; head(y)
X = data.frame(data[,5]); head(X)
x_name = colnames(data)[5]; x_name
colnames(X) = x_name; head(X)
factor_vars = X

#system.time({ out_list = anova_func(y, X, factor_vars) })

out_list

```

Dassit from me for now. Will elaborate when we speak next.

Sudhir
